Short-term synaptic plasticity and threshold adaptation act in synergy to give rise to deviance detection in a spiking model of neuronal culture

### 2022-12-05 paper-1/manuscript.tex
Adjusted tex for pandoc word export

\begin{figure*}%
    \centering
    \includegraphics{fig1.eps}
    \caption{\textbf{a} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from a small subset. \textbf{b} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to each of 5 stimulation sites, which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{c} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black, $\tau$ = 30 ms), firing threshold baseline (gray) and adaptive threshold (green, $\tau$ = 1 s) demonstrating TA. Middle: Excitatory ($g_e$, $\tau$ = 2 ms) and inhibitory ($g_i$, $\tau$ = 4 ms) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line)}
    \label{fig1}
\end{figure*}

@@

@@

@@

@@

@@

@@

@@

\bibliography{bibliography}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{\textbf{a} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from a small subset. \textbf{b} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to each of 5 stimulation sites, which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{c} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black, $\tau$ = 30 ms), firing threshold baseline (gray) and adaptive threshold (green, $\tau$ = 1 s) demonstrating TA. Middle: Excitatory ($g_e$, $\tau$ = 2 ms) and inhibitory ($g_i$, $\tau$ = 4 ms) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line)}
    \label{fig1}
\end{figure*}

\textbf{Fig. 1 a} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from a small subset. \textbf{b} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to each of 5 stimulation sites, which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{c} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black, $\tau$ = 30 ms), firing threshold baseline (gray) and adaptive threshold (green, $\tau$ = 1 s) demonstrating TA. Middle: Excitatory ($g_e$, $\tau$ = 2 ms) and inhibitory ($g_i$, $\tau$ = 4 ms) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line)

@@

\textbf{Fig. 2 a} Illustration of the oddball paradigm used. Each square represents a stimulus, with the target (A) in dark blue, and other stimuli in lighter colors. The temporal extent of stimulation is exaggerated; in reality, stimulation was delivered in a single time step, and the response allowed to develop freely for 500 ms. \textbf{b} Response of a neuronal cell culture in vitro, adapted from \cite{Kubota2021-dx}. \textbf{c} Model response, showing average network-wide spike counts after stimulation with target (A) in each sequence. \textbf{d} Mismatch, SSA and deviance detection indices across networks and stimulus pairs (n = 120). In this and later figures, notches indicate the confidence interval for the median, calculated by bootstrap with 10000 iterations. All indices were positive (mismatch: t = 11, p = 6.7e-21, SSA: t = 7.8, p = 1.05e-12, deviance detection: t = 6.5, p = 1.05e-09)

@@

\textbf{Fig. 3} Deviance detection and SSA indices under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. \textbf{b, c} Deviance detection and SSA indices across networks and stimulus pairs. Significance was assessed by 1-sample t-tests. TODO p-values: text or table?

@@

\textbf{Fig. 4} Responses and threshold adaptation in a TA-only ablation. \textbf{a, b, c} Post-stimulus histograms across target trials, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev and msc sequence, respectively; right column, contrast between these two. \textbf{d} Per-neuron relationship between the contrast (dev - msc) in $V_{TA}$ at the start of trials, and the contrast in average number of spikes fired, excluding inhibitory neurons, and associated linear regression (r = -0.424, p = 3.1e-36). \textbf{e} Contribution of $V_{TA}$ and $V_{syn}$ to bins with increased firing in dev (i.e., red bins in the spike probability contrast, \textbf{a}), weighted by that same contrast. \textbf{f} Contribution of $V_{TA}$ relative to the sum of the weighted contributions in \textbf{e}. \textbf{g} Relative contribution of $V_{TA}$ across networks, showing median (solid line) and inter-quartile range (shaded area)

@@

\textbf{Fig. 5 a} $V_{TA}$ at the start of A trials (mean over trials), with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence. \textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B/C/D/E in msc). \textbf{c} Left and middle: Replacement and adaptation components of the non-target response contrast (\textbf{b}, right column), where replacement is the contrast (B msc - C/D/E msc), and adaptation is the contrast (B dev - B msc). Right: Network-wide mean replacement and adaptation values across networks and target stimuli. The adaptation effect is less than zero (t = -9, p = 2.02e-15), and less than the replacement effect (t = 2.3, p = 0.0119), which is not different from zero (t = -0.69, p = 0.246). \textbf{d} Per-neuron relationship between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation, and the contrast in $V_{TA}$ at the start of target trials, excluding inhibitory neurons, and associated linear regression (r = 0.99, p = 0). \textbf{e} The same relationship as \textbf{d}, but using the medians across excitatory neurons, and plotting one point per network and target stimulus. The values are highly correlated (r = 0.956, p = 5.98e-65)

@@

\textbf{Fig. 6} Contrast between the ablated model with only TA (shown in Figure \ref{fig5}) and the full model with both STD and TA. \textbf{a}, \textbf{b} Data shown in Figure \ref{fig5}ab, dev and msc columns, subtracted from the corresponding data for the full model. \textbf{c} Equivalent contrast (full - ablated model) in the response to A, in columns as above. \textbf{d} Network-wide mean $V_{TA}$ before A trials across networks. Left: Ablated and full model averages. Right: Difference between these model conditions. With the introduction of STD, $V_{TA}$ decreased more in dev than in msc (t = -3.49, p = 0.000342). \textbf{e} Mean number of spikes per target trial across networks. Left: Ablated and full model averages. Right: Difference between these model conditions. With the introduction of STD, the response decreased more in the std sequence than msc (t = -1.9, p = 0.0297), and more in msc than in dev (t = -2.99, p = 0.00168)

@@

\textbf{Fig. 7 a} Estimated average STD-derived postsynaptic depression $D$ in target (A) trials (top row) and B trials (bottom row). Note that the difference between A and B in msc (middle column) is due only to the network configuration. \textbf{b} Statistics across networks of the contrast (dev - msc) in $D$ (cf. \textbf{a}, right column), and in $V_{TA}$ at the start of A trials (cf. Figure \ref{fig6}d). In each network, averages were taken over neurons with at least 0.2 spikes per dev trial (B trials for $\Delta D^B$, A trials for $\Delta D^A$ and $\Delta V_{TA}$). Whereas the third box ($\Delta V_{TA}$ total) indicates the averages in the full model, the final box ($\Delta V_{TA}$ added) indicates the difference between the full and the ablated (TA-only) model. This additional amount of $\Delta V_{TA}$, caused indirectly by the addition of STD, was less (i.e., more negative) than the directly STD-related $\Delta D^A$ itself (t = -2.09, p = 0.0195)

@@

\bibliography{refs}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl
